---
title: "R environments"
output:
   github_document:
      toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "img/")
suppressPackageStartupMessages(library(tidyverse))
```

## Introduction

Create a new environment to avoid namespace collisions.

```{r first_env}
first_env <- new.env()
source("code/hello.R", local = first_env)
first_env$hello()
```

Create another new environment.

```{r second_env}
second_env <- new.env()
source("code/hello2.R", local = second_env)
second_env$hello()
```

`.GlobalEnv` is the Global Environment which is the workspace environment where user-defined objects and functions are stored by default. It is the environment where all objects (variables, functions, etc.) created interactively in an R session are stored unless otherwise specified.

```{r global_env}
.GlobalEnv
```

Second environment.

```{r second_env_output}
second_env
```

If we list objects in `.GlobalEnv`, we will see nothing because the `hello()` functions are stored in different environments.

```{r list_global_env}
lsf.str(envir = .GlobalEnv)
```

List objects in `first_env`.

```{r list_first_env}
lsf.str(envir = first_env)
```

List objects in `second_env`.

```{r list_second_env}
lsf.str(envir = second_env)
```

## Session info

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```
