---
title: "Experimenting with geom_tile"
output:
   github_document:
      toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "img/")
library(ggplot2)
```

# Introduction

Making plots with [geom_tile](https://ggplot2.tidyverse.org/reference/geom_tile.html).

## Plots

Example data from the documentation.

```{r eg_data}
df <- data.frame(
  x = rep(c(2, 5, 7, 9, 12), 2),
  y = rep(c(1, 2), each = 5),
  z = factor(rep(1:5, each = 2)),
  w = rep(diff(c(0, 4, 6, 8, 10, 14)), 2)
)

df
```

Plot rectangles with default width.

```{r plot_rect}
my_interval <- 1
ggplot(df, aes(x, y)) +
  geom_tile(aes(fill = z), colour = "grey50") +
  scale_x_continuous(breaks = round(seq(0, max(df$x)+1, by = my_interval))) +
  scale_y_continuous(breaks = round(seq(0, max(df$y), by = my_interval)))
```

Plot rectangles with default width.

```{r plot_rect_width}
my_interval <- 1
ggplot(df, aes(x, y, width = w)) +
  geom_tile(aes(fill = z), colour = "grey50") +
  scale_x_continuous(breaks = round(seq(0, max(df$x)+1, by = my_interval))) +
  scale_y_continuous(breaks = round(seq(0, max(df$y), by = my_interval)))
```

To draw squares, use `coord_fixed()`; to create a border around each "cell", decrease the `width` and `height` slightly, and use `linewidth`.

```{r squares}
df2 <- data.frame(
  x = rep(seq(from = 1, to = 9, by = 2), 2),
  y = rep(c(1, 3), each = 5),
  z = factor(rep(1:2, each = 5)),
  ranking = 1:10
)

my_interval <- 1
ggplot(df2, aes(x, y)) +
  geom_tile(aes(fill = z, width = 1.9, height = 1.9), colour = "black", linewidth = 1) +
  coord_fixed() +
  theme_void() +
  theme(legend.position = "none")
```

Add text to each "cell" using `geom_text()`.

```{r add_text}
my_interval <- 1
ggplot(df2, aes(x, y)) +
  geom_tile(aes(fill = z, width = 1.9, height = 1.9), colour = "black", linewidth = 1) +
  geom_text(aes(x, y, label = ranking), size = 5) +
  coord_fixed() +
  theme_void() +
  theme(legend.position = "none")
```

Gradient colour using `scale_fill_gradient()`.

```{r scale_fill_gradient}
my_interval <- 1
ggplot(df2, aes(x, y)) +
  geom_tile(aes(fill = ranking, width = 1.9, height = 1.9), colour = "black", linewidth = 1) +
  geom_text(aes(x, y, label = ranking), size = 5) +
  coord_fixed() +
  theme_void() +
  theme(legend.position = "none") +
  scale_fill_gradient(low = "yellow", high = "red")
```

Use `RColorBrewer::display.brewer.all()` to find a nice palette and then use `RColorBrewer::brewer.pal()` to generate the colours. Change `ranking` into a factor and then use `scale_fill_discrete()` for discrete colouring.

```{r scale_fill_discrete}
n <- length(unique(df2$ranking))

my_interval <- 1
ggplot(df2, aes(x, y)) +
  geom_tile(aes(fill = factor(ranking), width = 1.9, height = 1.9), colour = "black", linewidth = 1) +
  geom_text(aes(x, y, label = ranking), size = 5) +
  coord_fixed() +
  theme_void() +
  theme(legend.position = "none") +
  scale_fill_discrete(RColorBrewer::brewer.pal(n, "Set3"))
```

## Session info

This document was generated by running `geom_tile.Rmd` in RStudio Server.

```{r time, echo=FALSE}
Sys.time()
```

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```