---
title: "Installing R packages"
output:
   github_document:
      toc: true
---

# Notes on installing R packages

The function `install.packages` from the {utils} packages can install packages from repositories or local files.

Packages are installed to `.libPaths` by default.

```{r lib_paths}
.libPaths()
```

Install the dependencies required to render this document.

```{r install_dep}
my_packages <- c("rmarkdown", "knitr")

sapply(my_packages, function(x){
  if(!require(x, character.only = TRUE)){
    install.packages(x, quiet = TRUE)
  }
})
```

Use a specific [mirror](https://cran.r-project.org/mirrors.html).

```{r install_remotes}
if(!require(remotes)){
  install.packages("remotes", repos = "https://cran.ism.ac.jp/", quiet = TRUE)
}
packageVersion("remotes")
```

Use the {remotes} package to install a specific version of a package. If the package is available on CRAN, look at the "Old sources:" section to find older version.

```{r install_beepr_version}
remotes::install_version("beepr", version = "1.1", repos = "https://cran.ism.ac.jp/", quiet = TRUE)
packageVersion("beepr")
```

Update {beepr} by using `install.packages`.

```{r update_beepr}
install.packages("beepr", repos = "https://cran.ism.ac.jp/", quiet = TRUE)
packageVersion("beepr")
```

## Session info

This document was generated by rendering `installing_packages.Rmd` in the RStudio Server instance created by `start_rstudio.sh`.

```{r time, echo=FALSE}
Sys.time()
```

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```
