---
title: "Iteration"
output:
   github_document:
      toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "img/")
library(tidyverse)
```

# Iterating using {dplyr} and {purrr}

Ways to iterate using packages from the Tidyverse.

## Working with multiple columns

Use `across()`

```{r across}
mtcars |>
  summarise(
    n = n(),
    across(mpg:qsec, median)
  )
```

Multiple functions by using a list (and modifying column names by using `.names`).

```{r across_multiple_functions}
mtcars |>
  summarise(
    across(
      mpg:qsec,
      list(
        mean = mean,
        median = median
      ),
      .names = "{.fn}_{.col}"
    )
  )
```

Use `where()` to filter columns.

```{r across_where}
ChickWeight |>
  summarise(
    across(where(is.double), mean)
  )
```

## Pivot longer

Compare `across` with `pivot_longer`.

```{r across_again}
mtcars |>
  summarise(
    across(mpg:qsec, median)
  )
```

Using `pivot_longer` is useful if `across` isn't flexible enough.

```{r pivot_longer}
mtcars |>
  pivot_longer(mpg:qsec) |>
  group_by(name) |>
  summarise(median = median(value))
```

## Reading multiple files

[Create multiple files](https://davetang.org/muse/2023/06/23/reading-list-of-files-into-single-r-data-frame/).

```{r random_files}
random_df <- function(num_row = 100, num_col = 100, seed = 1984){
  set.seed(seed)
  matrix(
    data = 
      runif(
        n = num_row * num_col,
        min = 0,
        max = 1
      ),
    nrow = num_row
  ) |> as.data.frame()
}

random_files <- function(nfiles, prefix = 'x', outdir = 'random', leading_zero = 6){
  if(!dir.exists(outdir)){
    dir.create(outdir)
  }
  purrr::map(1:nfiles, function(x){
    write.csv(
      x = random_df(seed = x),
      file = paste0(outdir, '/', prefix, stringr::str_pad(x, leading_zero, pad = 0), ".csv"),
      row.names = FALSE
    )
  }) -> dev_null
}

random_files(5)

list.files("random")
```

Read the files using `list_rbind`; remember to use `set_names` first and the `names_to` parameter.

```{r list_rbind}
list.files(path = "random", pattern = "csv$", full.names = TRUE) |>
  purrr::set_names(basename) |>
  purrr::map(readr::read_csv, show_col_types = FALSE) |>
  purrr::list_rbind(names_to = "file") -> my_df

dim(my_df)
```

Take a look at the data frame.

```{r head_my_df}
head(my_df)
```

## Session info

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```
