---
title: "Object-Oriented Programming in R"
output:
   github_document:
      toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "img/")

.libPaths('/packages')
my_packages <- c('sloop')

using<-function(...) {
   # https://stackoverflow.com/a/44660688
   libs<-unlist(list(...))
   req<-unlist(lapply(libs,require,character.only=TRUE))
   need<-libs[req==FALSE]
   if(length(need)>0){
      install.packages(need)
      lapply(need,require,character.only=TRUE)
   }
}

using(my_packages)
```

# Introduction

Notes from the [Advanced R](https://adv-r.hadley.nz/oo.html).

Everything that exists in R is an object but this does not mean that everything
is object-oriented. In R there are **base objects** and **OO objects** and we
can use `is.object()` or `sloop::otype()` to tell which is which.

```{r object_type}
is.object(1:10)
sloop::otype(1:10)

is.object(mtcars)
sloop::otype(mtcars)
```

Technically, the difference between base and OO objects is that OO objects have
a "class" attribute.

```{r attribute}
attr(1:10, "class")
attr(mtcars, "class")
```

The `class()` function is safe to apply to S3 and S4 objects but returns
misleading results when applied to base objects. It is safer to use
`sloop::s3_class()`, which returns the implicit class that the S3 and S4 systems
will use to pick methods.

```{r s3_class}
x <- matrix(1:4, nrow = 2)
class(x)
sloop::s3_class(x)
```

## Base types

While only OO objects have a class attribute, every object has a **base type**.

```{r typeof}
typeof(1:10)
typeof(mtcars)
```

Base types do not form an OOP system because functions that behave differently
for different base types are primarily written in C code that uses switch
statements. This means that only R-core can create new types, and creating a new
type is a lot of work because every switch statement needs to be modified to
handle a new case.

### Numeric type

R uses numeric to mean three slightly different things:

1. In some places numeric is used as an alias for the double type.
2. In the S3 and S4 systems, numeric is used as a shorthand for either integer
   or double type, and is used when picking methods:

```{r numeric_type}
sloop::s3_class(1)
sloop::s3_class(1L)
```

3. `is.numeric()` tests for objects that _behave_ like numbers. For example,
   factors have type "integer" but do not behave like numbers.

```{r factor_type}
typeof(factor("x"))
is.numeric(factor("x"))
```

## S3

An S3 object is a base type with at least a `class` attribute. As an example, consider the factor. Its base type is the integer vector and it has a `class` attribute of "factor" and a `levels` attribute that stores the possible levels.

```{r factor_example}
f <- factor(c("a", "b", "c"))

typeof(f)
attributes(f)
```

You can get the underlying base type by using `unclass()`, which strips the class attribute, causing it to lose its special behaviour.

```{r unclass_a_factor}
unclass(f)
```

An S3 object behaves differently from its underlying base type whenever it is passed to a **generic** (short for generic function). The easiest way to tell if a function is a generic is to use `sloop::ftype()` and look for "generic" in the output.

```{r ftype}
ftype(print)
ftype(str)
ftype(unclass)
```

A generic function defines an interface, which uses a different implementation depending on the class of an argument (almost always the first argument).

```{r print_unclass}
print(f)

# stripping class reverts to integer behaviour
print(unclass(f))
```

The generic is a middleman: its job is to define the interface (i.e. the arguments) then find the right implementation for the job. The implementation for a specific class is called a **method**, and the generic finds that method by performing **method dispatch**.

You can use `sloop::s3_dispatch()` to see the process of method dispatch.

```{r s3_dispath}
sloop::s3_dispatch(print(f))
```

Note that S3 methods are functions with a special naming scheme `generic.class()`. For example, the `factor` method for the `print()` generic is called `print.factor()`. **You should never call the method directly, but instead rely on the generic to find it for you**.

Generally, you can identify a method by the presence of `.` in the function name but use `sloop::ftype()` to confirm.

```{r ftype_check}
ftype(t.test)
ftype(t.data.frame)
```

**You can not see the source code for most S3 methods by typing their names unlike most functions. That's because S3 methods are not usually exported: they live only inside the package, and are not available from the global environment**. Use `sloop::s3_get_method()` to get the method code.

```{r s3_get_method}
sloop::s3_get_method(weighted.mean.Date)
```

## Classes

S3 has no formal definition of a class: to make an object an instance of a class, you simply **set the class attribute**. You can do that during creation with `structure()`, or after with `class<-()`.

```{r structure_class}
# Create and assign class in one step
x <- structure(list(), class = "my_class")

# Create, then set class
x <- list()
class(x) <- "my_class"
```

You can determine the class of an S3 object with `class(x)` and see if an object is an instance of a class using `inherits(x, "classname")`.

```{r class_inherits}
class(x)
inherits(x, "my_class")
inherits(x, "your_class")
```

The class name can be any string, but Hadley recommends using only letters and `_`. Avoid `.` because it can be confused with the `.` separator between a generic name and a class name. When using a class in a package, it is recommended that the package name be included with the class name. That ensures you won't accidentally clash with a class defined by another package.

S3 has no checks for correctness which means you can change the class of existing objects.

```{r change_class_of_existing_obj}
mod <- lm(log(mpg) ~ log(disp), data = mtcars)
class(mod)

class(mod) <- "Date"
class(mod)
```

When creating a class, it is recommended that these three functions are provided:

* A low-level **constructor**, `new_myclass()`, that efficiently creates new objects with the correct structure.
* A **validator**, `validate_myclass()`, that performs more computationally expensive checks to ensure that the object has correct values.
* A user-friendly **helper**, `myclass()`, that provides a convenient way for others to create objects of your class.

For simple classes a validator is not necessary and the helper can be skipped if the class is for internal use only, but a constructors should always be provided.

### Constructors

S3 doesn't provide a formal definition of a class, so it has no built-in way to ensure that all objects of a given class have the same structure (i.e. the same base type and the same attributes with the same types). Instead, you must enforce a consistent structure by using a **constructor**.

The constructor should follow three principles:

1. Be called `new_myclass()`.
2. Have one argument for the base object, and one for each attribute.
3. Check the type of the base object and the types of each attribute.

We'll get started with a constructor for the simplest S3 class: `Date`. A `Date` is just a double with a single attribute: its `class` is "Date".

```{r new_date}
new_Date <- function(x = double()) {
   stopifnot(is.double(x))
   structure(x, class = "Date")
}

new_Date(c(-1, 0, 1))
```

The purpose of constructors is to help you, the developer. That means you can keep them simple, and you don't need to optimise error messages for public consumption. If you expect users to also create objects, you should create a friendly helper function, called `class_name`.

A slightly more complicated constructor is that for `difftime`, which is used to represent time differences. It is again built on a double, but has a `units` attribute that must take one of a small set of values.

```{r new_difftime}
new_difftime <- function(x = double(), units = "secs"){
  stopifnot(is.double(x))
  units <- match.arg(units, c("secs", "mins", "hours", "days", "weeks"))
  
  structure(
    x,
    class = "difftime",
    units = units
  )
}

new_difftime(c(1, 10, 3600), "secs")
new_difftime(52, "weeks")
```

The constructor is a developer function: it will be called in many places, by an experienced user. That means it is OK to trade a little safety in return for performance, and you should avoid potentially time-consuming checks in the constructor.

### Validators

More complicated classes require more complicated checks for validity. Take factors, for example. A constructor only checks that types are correct, making it possible to create malformed factors.

```{r new_factor, error=TRUE}
new_factor <- function(x = integer(), levels = character()){
  stopifnot(is.integer(x))
  stopifnot(is.character(levels))
  
  structure(
    x,
    levels = levels,
    class = "factor"
  )
}

new_factor(1:5, "a")
```

Rather than encumbering the constructor with complicated checks, it's better to put them in a separate function. Doing so allows you to cheaply create new objects when you know that the values are correct, and easily re-use the checks in other places.

```{r validate_factor, error=TRUE}
validate_factor <- function(x){
  values <- unclass(x)
  levels <- attr(x, "levels")
  
  if (!all(!is.na(values) & values > 0)){
    stop(
      "All `x` values must be non-missing and greater than zero",
      call. = FALSE
    )
  }
  
  if (length(levels) < max(values)){
    stop(
      "There must be at least as many `levels` as possible values in `x`",
      call. = FALSE
    )
  }
  
  x
}

validate_factor(new_factor(1:5, "a"))
validate_factor(new_factor(0:1, "a"))
```

The validator function is called primarily for its side-effects (throwing an error if the object is invalid) so you'd expect it to invisibly return its primary input. However, it is useful for validation methods to return visibly too.

## Session info

This document was generated by rendering `oop.Rmd` using `rmd_to_md.sh`.

```{r time, echo=FALSE}
Sys.time()
```

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```
