---
title: "Writing R packages"
output:
   github_document:
      toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "img/")
library(ggplot2)
```

# Introduction

Notes from the [R Packages book](https://r-pkgs.org/).

## The Big Picture

From [The Whole Game chapter](https://r-pkgs.org/whole-game.html).

* Use the `devtools` package, which contains a set of packages that support package development.
* Use `create_package()` to initialise a new package.
* Use `use_r(function_name)` to create and/or open a script in `R/`.
* Use `document()` to build the documentation.
* Use `load_all()` to load all functions in `R/`; `load_all()` simulates the process of building, installing, and attaching.
* Use `check()` to check that everything is working.
* Use `install()` to install your package.
* Use `use_git()` to make a Git repository.
* Use `use_github()` to connect to GitHub.
* Use `use_readme_rmd()` to create a template `README.Rmd` file then use `build_readme()` to render `README.md`.

## Session info

This document was generated by rendering `packages.Rmd` in RStudio Server.

```{r time, echo=FALSE}
Sys.time()
```

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```
