---
title: "Vectors"
output:
   github_document:
      toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path = "img/")
library(tidyverse)
```

# Introduction

The most important family of data types in base R are vectors (from Advanced R). All the other data types are known as "node" types, which include things like functions and environments. You are most likely to come across this highly technical term when using `gc()`: the "N" in `Ncells` stands for nodes and the "V" in `Vcells` stands for vectors.

Vectors come in two flavours:

1. atomic vectors and
2. lists.

They differ in terms of their elements' types: for atomic vectors, all elements must have the same type; for lists, elements can have different types. While not a vector, `NULL` is closely related to vectors and often serves the role of a generic zero length vector.

Every vector can also have **attributes**, which you can think of as a named list of arbitrary metadata. Two attributes are particularly important:

1. the **dimension** attribute turns vectors into matrices and arrays and
2. the **class** attribute powers the S3 object system.

## Atomic vectors

There are four primary types of atomic vectors:

1. logical,
2. integer,
3. double, and
4. character (which contains strings).

Collectively integer and double vectors are known as numeric vectors. There are two rare types: complex and raw.

### Scalars

Each of the four primary types has a special syntax to create an individual value, AKA a scalar (a vector of length one):

* Logicals can be written in full (`TRUE` or `FALSE`) or abbreviated (`T` or `F`)
* Doubles can be specified in decimal, scientific, or hexadecimal form. There are three special values unique to doubles: `Inf`, `-Inf`, and `NaN` (not a number). These are special values defined by the floating point standard.
* Integers are written similarly to doubles but must be followed by `L` (Long integer) and can not contain fractional values
* Strings are surrounded by `"` or `'` and special characters are escaped with `\`.

You can determine the type of a vector with `typeof()` and its length with `length()`.

### Missing values

R represents missing or unknown values with the special sentinel value `NA`. Missing values tend to be infectious as most computations involving a missing value will return another missing values. There are only a few exceptions to this rule and these occur when some identity holds for all possible inputs.

Propagation ofmissingness leads to a common mistake when determining which values in a vector are missing.

```{r}
x <- c(NA, 5, NA, 10)
x == NA
```

The result is correct because there's no reason to believe that **one missing value has the same value as another**, i.e., since `NA` is an unknown value, we can't subset using `NA` and if we do everything is unknown.

For testing if there are `NA`'s, use `is.na()`.

```{r}
x <- c(NA, 5, NA, 10)
is.na(x)
```

### Testing and coercion

Generally, you can **test** if a vector is of a given type with an `is.*()` function, but these functions need to be used with care.

`is.logical()`, `is.integer()`, `is.double()`, and `is.character()` do what you might expect, which is they test if a vector is a logical, integer, double, or character.

Avoid `is.vector()`, `is.atomic()`, and `is.numeric()` because they don't test if you have a vector, atomic vector, or numeric vector. You will need to read the documentation carefully to figure out what they actually do.

For atomic vectors, type is a property of the entire vector: all elements must be the same type. When you attempt to combine different types they will be **coerced** in a fixed order:

character -> double -> integer -> logical

For example, combining a character and an integer yields a character:

```{r}
str(c("a", 1L))
```

Coercion often happens automatically and most mathematical functions (`+`, `log`, `abs`, etc.) will coerce to numeric. This coercion is particularly useful for logical vectors because `TRUE` becomes and `FALSE` becomes 0.

Generally, you can deliberately coerce by using an `as.*()` function, like `as.logical()`, `as.integer()`, `as.double()`, or `as.character()`. Failed coercion of strings generates a warning and a missing value.

```{r}
as.integer(c("1", "1.5", "a"))
```

## Attributes

You might have noticed that the set of atomic vectors does not include a number of important data structures like matrices, arrays, factors, or datetimes. These types are built on top of atomic vectors by adding attributes.

### Getting and setting

You can think of attributes as name-value pairs that attach metadata to an object. Attributes behave like named lists but are actually pairlists; pairlists are functionally indistinguishable from lists but are profoundly different under the hood.

Individual attributes can be:

1. retrieved and modified with `attr()`,

```{r}
a <- 1:3
# attr(x, which) <- value
# where which is a non-empty character string specifying which attribute is to be accessed.
attr(a, "x") <- "abcdef"
attr(a, "x")
```

2. retrieved en masse with `attributes()`, and

```{r}
attr(a, "y") <- 4:6
str(attributes(a))
```

3. set en masse with `structure()`.

```{r}
a <- structure(
  1:3,
  x = "abcdef",
  y = 4:6
)

a
class(a)
str(attributes(a))
```

Attributes should generally be thought of as ephemeral. For example, most attributes are lost by most operations.

```{r}
attributes(a[1])
```

There are only two attributes that are routinely preserved:

1. **names** - a character vector giving each element a name
2. **dim** - short for dimensions, an integer vector, used to turn vectors into matrices or arrays

To preserve other attributes, you will need to create your own S3 class.

### Names

You can name a vector in three ways:

1. when creating it:

```{r}
x <- c(a = 1, b = 2, c = 3)
x
```

2. by assigning a character vector to `names()`:

```{r}
x <- 1:3
names(x) <- c('a', 'b', 'c')
x
```

3. inline with `setNames()`:

```{r}
x <- setNames(1:3, c('a', 'b', 'c'))
x
```

Avoid using `attr(x, "names")` as it requires more typing and is less readable than `names(x)`.

```{r}
x <- 1:3
attr(x, 'names') <- c('a', 'b', 'c')
x
```

You can remove names from a vector by using `unname(x)` or `names(x) <- NULL`.

To be useful with character subsetting, names should be unique and non-missing, but this is not enforced by R. Depending on how the names are set, missing names may be either "" or `NA_character_`. If all names are missing, `names()` will return `NULL`.

### Dimensions

Adding a `dim` attribute to a vector allows it to behave like a 2-dimensional **matrix** or a multi-dimensional **array**. Matrices and arrays are primarily mathematical and statistical tools, not programming tools. Their most important feature is multidimensional subsetting.

You can create matrices and arrays with `matrix()` and `array()`, or by using the assignment form of `dim()`.

Using `matrix()` and two scalar arguments that specify the row and column sizes

```{r}
a <- matrix(1:6, nrow = 2, ncol = 3)
a
class(a)
```

Using a vector argument to describe all dimensions.

```{r}
b <- array(1:12, c(2, 3, 4))
b
```

Modify an object in place by setting `dim()`:

```{r}
d <- 1:6
dim(d) <- c(3, 2)
d
class(d)
```

Many of the functions for working with vectors have generalisations for matrices and arrays:

| Vector | Matrix | Array |
| ------ | ------ | ----- |
| `names()` | `rownames()`, `colnames()` | `dimnames()` |
| `length()` | `nrow()`, `ncol()` | `dim()` |
| `c()` | `rbind()`, `cbind()` | `abind::abind()` |
| - | `t()` | `aperm()` |
| `is.null(dim(x))` | `is.matrix()` | `is.array()` |

A vector without a `dim` attribute set is often thought of as 1-dimensional, but actually has `NULL` dimensions. You can also have matrices with a single row or column, or arrays with a single dimension. They may print similarly, but will behave differently. The differences aren't too important but it's useful to know they exist in case you get strange output from a function. As always, use `str()` to reveal the differences.

## Session info

Session info.

```{r session_info, echo=FALSE}
sessionInfo()
```
