---
title: "R package workshop"
format:
  gfm:
    toc: true
date: '`r Sys.Date()`'
date-format: iso
engine: knitr
---

```{r load_libraries, include=FALSE}
library(tidyverse)
```

# TL;DR

Install packages, check whether the system is ready to build packages, initialise package, add license, create file to hold functions, and check for issues.

```{r}
#| label: tl_dr
#| eval: false

pkgs <- c("devtools", "usethis", "roxygen2", "testthat", "knitr", "ggplot2", "rlang")
install.packages(pkgs, Ncpus = 2)
pkgbuild::check_build_tools()

usethis::create_package(path = '/tmp/mypkg')
usethis::use_mit_license("Dave Tang")
usethis::use_r("colours")
devtools::document()
usethis::use_testthat()
usethis::use_test("colours")
devtools::test()
devtools::check()
```

# Introduction

This [R package workshop](https://combine-australia.github.io/r-pkg-dev/) was created by COMBINE, an association for Australian students in bioinformatics, computational biology, and related fields. It is licensed under a [Creative Commons Attribution 4.0 International (CC BY 4.0) license](https://creativecommons.org/licenses/by/4.0/).

An R package is a collection of functions that are bundled together in a way that makes it easy to share. Usually these functions are designed to work together to complete a specific task such as analysing a particular kind of data. Packages are the best way to distribute code and documentation; even if you never intend to share your package it is useful to have a place to store your commonly used functions. It's a bit more effort now but it will save you a lot of time in the long run.

This workshop teaches a modern package development workflow that makes use of packages designed to help with writing packages. The two main packages are {devtools} and {usethis}:

* {devtools} contains functions that will help with development tasks such as checking, building and installing packages.
* {usethis} contains a range of templates and handy functions for making life easier, many of which were originally in {devtools}.

Other packages used in this workshop include:

* {roxygen2} for function documentation
* {testthat} for writing unit tests
* {knitr} for building vignettes

Install the required packages for this workshop.

```{r}
#| label: required_packages
#| eval: false

pkgs <- c("devtools", "usethis", "roxygen2", "testthat", "knitr", "ggplot2", "rlang")
install.packages(pkgs, Ncpus = 2)
```

Some stages of the package development process can require other programs to be installed on your computer. To check that you have everything you need run the following:

```{r}
#| label: check_build_tools

pkgbuild::check_build_tools()
```

# Getting started

To begin building a package, we need a package name; names can only consist of letters, numbers and dots (.) and must start with a letter. While all of these are allowed it is generally best to stick to just lowercase letters. Having a mix of lower and upper case letters can be hard for users to remember (is it RColorBrewer or Rcolorbrewer or rcolorbrewer?). For this workshop, we are going to use {mypkg}.

To create a template for our package, use `usethis::create_package()`.

```{r}
#| label: create_package
#| eval: false
 
usethis::create_package(path = '/tmp/mypkg')
```

Several files get created.

```{bash}
ls -R /tmp/mypkg
```

* `DESCRIPTION` contains the metadata for your package.
* `NAMESPACE` describes the functions in our package; it should not be manually edited.
* `mypkg.Rproj` is a RStudio project file.
* `R` is the directory that will hold all our R code.

These files are the minimal amount that is required for a package but we will create other files as we go along. Some other useful (hidden) files have also been created by {usethis}:

* `.gitignore` is useful if you use `git` for version control.
* `.Rbuildignore` is used to mark files that are in the directory but aren’t really part of the package and shouldn't be included when we build it. Most of the time you won't need to worry about this as {usethis} will edit it for you.

## DESCRIPTION

The `DESCRIPTION` file is one of the most important parts of a package. It contains all the metadata about the package, things like what the package is called, what version it is, a description, who the authors are, what other packages it depends on, etc.

```{bash}
cat /tmp/mypkg/DESCRIPTION
```

* The `Title` should be a single line in Title Case that explains what your package is.
* The `Description` is a paragraph which goes into a bit more detail about what the package does.

The `Authors@R` field shows an example of how to define an author. You can see that the example person has been assigned the author ("aut") and creator ("cre") roles. There must be at least one author and one creator for every package (they can be the same person) and the creator must have an email address. Other roles include:

* `cre`: the creator or maintainer of the package, the person who should be contacted when there are problems
* `aut`: authors, people who have made significant contributions to the package
* `ctb`: contributors, people who have made smaller contributions
* `cph`: copyright holder, useful if this is someone other than the creator (such as their employer)

## License

The software license describes how our code can be used and without one people must assume that it can't be used at all! For this example we will use the MIT license which basically says the code can be used for any purpose and doesn't come with any warranties. There are templates for some of the most common licenses included in {usethis}.

```{r}
#| label: use_mit_license
#| eval: false
 
usethis::use_mit_license("Dave Tang")
```

## Functions

Use `usethis::use_r()` to create an empty file in `R` to hold our function about colours.

```{r}
#| label: use_r
#| eval: false

usethis::use_r("colours")
```

There are no rules about how to organise your functions into different files but in general similar functions should be grouped together into the same file with a clear name. Putting all functions into a single file is not ideal but neither is having a separate file for each function. A good rule of thumb is that if you are finding it hard to locate a function you might need to move it to a new file.

The example for this workshop is a function that takes the red, green and blue values for a colour and returns a given number of shades. Save the function below into `R/colours.R`.

```{r}
#| label: make_shades

make_shades <- function(colour, n, lighter = TRUE) {
  # Convert the colour to RGB
  colour_rgb <- grDevices::col2rgb(colour)[, 1]
  
  # Decide if we are heading towards white or black
  if (lighter) {
    end <- 255
  } else {
    end <- 0
  }
  
  # Calculate the red, green and blue for the shades
  # we calculate one extra point to avoid pure white/black
  red   <- seq(colour_rgb[1], end, length.out = n + 1)[1:n]
  green <- seq(colour_rgb[2], end, length.out = n + 1)[1:n]
  blue  <- seq(colour_rgb[3], end, length.out = n + 1)[1:n]
  
  # Convert the RGB values to hex codes
  shades <- grDevices::rgb(red, green, blue, maxColorValue = 255)
  
  return(shades)
}
```

Usually when we write a new function we load it by copying the code to the console or sourcing the R file. When we are developing a package we want to try and keep our environment empty so that we can be sure we are only working with objects inside the package. Instead we can load functions using `devtools::load_all()`.

## Checking your package

The `devtools::check()` function runs a series of checks to make sure your package is working correctly. It is highly recommended that you run `devtools::check()` often and follow its advice to fix any problems.

```{r}
#| label: devtools_check
#| eval: false

devtools::check()
```

You will see all the different types of checks that {devtools} has run but the most important section is at the end where it tells you how many errors, warnings, and notes there are.

* Errors happen when your code has broken and failed one of the checks. If errors are not fixed your package will not work correctly.
* Warnings are slightly less serious but should also be addressed. Your package will probably work without fixing them but it is highly advised that you do.
* Notes are advice rather than problems. It can be up to you whether or not to address them but there is usually a good reason to.

Often the failed checks come with hints about how to fix them but sometimes they can be hard to understand.

## Documenting our function

The {roxygen2} package makes it easy to document our function by writing some special comments at the start of our function. This has the extra advantage of keeping the documentation with the code which make it easier to keep it up to date.

If you are using RStudio, open the "Code" menu and select "Insert Roxygen skeleton", which will create the template shown below. The parameters were automatically generated.

```{r}
#| label: roxygen_skeleton

#' Title
#'
#' @param colour 
#' @param n 
#' @param lighter 
#'
#' @return
#' @export
#'
#' @examples
```

Roxygen comments all start with `#'`. The first line is the title of the function followed by a blank line. Next is a paragraph that provides a more detailed description of the function.

The next section describes the parameters for the function marked by the `@param` field.

The next field is `@return`, which is used to describe what the function returns. This is usually fairly short but you should provide enough detail to make sure that the user knows what they are getting back.

The next field is `@export` and is a bit different because it doesn't add documentation to the help file but instead modifies the `NAMESPACE` file. Adding `@export` tells {Roxygen} that this is a function that we want to be available to the user. When we build the documentation, {Roxygen} will then add the correct information to the `NAMESPACE` file. If we had an internal function that wasn't meant to be used by the user we would leave out `@export`.

The last field in the skeleton is `@examples` and is where we put some short examples showing how the function can be used. These will be placed in the help file and can be run using `example("function")`.

```{r}
#| label: roxygen_skeleton_filled

#' Make shades
#'
#' Given a colour make n lighter or darker shades
#' 
#' @param colour The colour to make shades of
#' @param n The number of shades to make
#' @param lighter Whether to make lighter (TRUE) or darker (FALSE) shades
#'
#' @return A vector of n colour hex codes
#' @export
#'
#' @examples
#' # Five lighter shades
#' make_shades("goldenrod", 5)
#' # Five darker shades
#' make_shades("goldenrod", 5, lighter = FALSE)
```

There are other useful fields such as `@author` (specify the function author), `@references` (any associated references) and `@seealso` (links to related functions).

Use `devtools::document()` to build the documentation.

```{r}
#| label: document
#| eval: false

devtools::document()
```

## Testing

Unit tests are checks to make sure that a function works in the way that we expect. The examples in the documentation act like informal unit tests because they are run as part of the checking process but it is better to have something more rigorous.

One approach to writing unit tests is what is known as "test driven development". The idea is to write the tests before you write a function. This way you know exactly what a function is supposed to do and what problems there might be. While this is a good principal it can take a lot of advance planning.

A more common approach could be called "bug-driven testing". For this approach whenever we come across a bug we write a test for it before we fix it, that way the same bug should never happen a again. When combined with some tests for obvious problems this is a good compromise between testing for every possible outcome and not testing at all.

For example, let's see what happens when we ask `make_shades()` for a negative number of shades.

```{r}
#| label: make_shades_error
#| error: true

make_shades("goldenrod", -1)
```

Before we make any changes to the function let's design some tests to make sure we get what we expect. There are a few ways to write unit tests for R packages but we are going to use the {testthat} package. We can set everything up with {usethis}.

```{r}
#| label: use_test_that
#| eval: false

usethis::use_testthat()
```

A `tests/` directory will be created to hold all our tests. There is also a `tests/testthat.R` file which looks like this:

```{r}
#| label: test_that_r_script
#| eval: false

# This file is part of the standard setup for testthat.
# It is recommended that you do not modify it.
#
# Where should you do additional test configuration?
# Learn more about the roles of various files in:
# * https://r-pkgs.org/testing-design.html#sec-tests-files-overview
# * https://testthat.r-lib.org/articles/special-files.html

library(testthat)
library(mypkg)

test_check("mypkg")
```

The above script makes sure that our tests are run with `devtools::check()`. To open a new test file use `usethis::use_test()`.

```{r}
#| label: use_test_colours
#| eval: false

usethis::use_test("colours")
```

`use_test("colours")` creates `tests/testthat/test-colours.R`, which matches up with the R files but with `test-` as the prefix. This test file comes with a small example that shows how to use {testthat}.

```{r}
#| label: test_that_example
#| eval: false

test_that("multiplication works", {
  expect_equal(2 * 2, 4)
})
```

Each set of tests starts with the `test_that()` function. This function has two arguments, a description and the code with the tests that we want to run. Inside the code section we see an expect function. This function also has two parts, the thing we want to test and what we expect it to be. There are different functions for different types of expectations.

For our test we want to use the `expect_error()` function, because that is what we expect.

```{r}
#| label: test_that_n_is_at_least_one
#| eval: false

test_that("n is at least 1", {
  expect_error(make_shades("goldenrod", -1),
               "n must be at least 1")
  expect_error(make_shades("goldenrod", 0),
               "n must be at least 1")
})
```

To run our tests we use `devtools::test()`. We will see that both of our tests failed. The first test fails because the error message is wrong and the second one because there is no error. Now that we have some tests and we know they check the right things we can modify our function to check the value of `n` and give the correct error.

Let's add some code to check the value of n. We will update the documentation as well so the user knows what values can be used.

```{r}
#| label: make_shades_check_n

#' Make shades
#'
#' Given a colour make \code{n} lighter or darker shades
#'
#' @param colour The colour to make shades of
#' @param n The number of shades to make, at least 1
#' @param lighter Whether to make lighter (\code{TRUE}) or darker (\code{FALSE})
#' shades
#'
#' @return A vector of \code{n} colour hex codes
#' @export
#'
#' @examples
#' # Five lighter shades
#' make_shades("goldenrod", 5)
#' # Five darker shades
#' make_shades("goldenrod", 5, lighter = FALSE)
make_shades <- function(colour, n, lighter = TRUE) {
  
  # Check the value of n
  if (n < 1) {
    stop("n must be at least 1")
  }
  
  # Convert the colour to RGB
  colour_rgb <- grDevices::col2rgb(colour)[, 1]
  
  # Decide if we are heading towards white or black
  if (lighter) {
    end <- 255
  } else {
    end <- 0
  }
  
  # Calculate the red, green and blue for the shades
  # we calculate one extra point to avoid pure white/black
  red <- seq(colour_rgb[1], end, length.out = n + 1)[1:n]
  green <- seq(colour_rgb[2], end, length.out = n + 1)[1:n]
  blue <- seq(colour_rgb[3], end, length.out = n + 1)[1:n]
  
  # Convert the RGB values to hex codes
  shades <- grDevices::rgb(red, green, blue, maxColorValue = 255)
  
  return(shades)
}
```

These kinds of checks for parameter inputs are an important part of a function that is going to be used by other people (or future you). They make sure that all the input is correct before the function tries to do anything and avoids confusing error messages. However they can be fiddly and repetitive to write. If you find yourself writing lots of these checks two packages that can make life easier by providing functions to do it for you are {checkmate} and {assertthat}.

In the code above we have used the `stop()` function to raise an error. If we wanted to give a warning we would use `warning()` and if just wanted to give some information to the user we would use `message()`. Using `message()` instead of `print()` or `cat()` is important because it means the user can hide the messages using `suppressMessages()` (or `suppressWarnings()` for warnings). Now if we try our tests again they should pass.

If you want to see what parts of your code need testing you can run the `devtools::test_coverage()` function (you will need to install the {DT} and {covr} packages first). This function uses the {covr} package to make a report showing which lines of your code are covered by tests.

## Dependencies

The `make_shades()` function produces shades of a colour but it would be good to see what those look like. Below is a new function called `plot_colours()` that can visualise them for us using {ggplot2}. Add the following function to `colours.R`.

```{r}
#| label: plot_colours
#| eval: false

#' Plot colours
#'
#' Plot a vector of colours to see what they look like
#' 
#' @param colours Vector of colour to plot
#'
#' @return A ggplot2 object
#' @export
#'
#' @examples
#' shades <- make_shades("goldenrod", 5)
#' plot_colours(shades)
plot_colours <- function(colours) {
  plot_data <- data.frame(Colour = colours)
  
  ggplot(plot_data,
         aes(x = .data$Colour, y = 1, fill = .data$Colour,
             label = .data$Colour)) +
    geom_tile() +
    geom_text(angle = "90") +
    scale_fill_identity() +
    theme_void()
}
```

If we run `devtools::check()`, we'll get the error:

```
Error in ggplot(plot_data, aes(x = .data$Colour, y = 1, fill = .data$Colour,  : 
    could not find function "ggplot"
```

Just like when we export a function in our package we need to make it clear when we are using functions in another package. To do this we can use `usethis::use_package()`.

```{r}
#| label: use_package
#| eval: false

usethis::use_package("ggplot2")
```

This will add {ggplot2} to the `Imports` field in `DESCRIPTION`. If we look in `DESCRIPTION` we will see the following line:

```
Imports:
    ggplot2
```

Those two lines tell us that our package uses functions in {ggplot2}. There are three main types of dependencies.

1. `Imports` is the most common. This means that we use functions from these packages and they must be installed when our package is installed.
2. The next most common is `Suggests`. These are packages that we use in developing our package (such as {testthat} which is already listed here) or packages that provide some additional, optional functionality. Suggested packages aren't usually installed so we need to do a check before we use them. The output of `usethis::use_package()` will give you an example if you add a suggested package.
3. The third type of dependency is `Depends`. If you depend on a package it will be loaded whenever your package is loaded. Depends shouldn't usually be used unless your package closely complements another package. An example of when Depends is necessary is if you package mainly operates on an object defined by another package.

Should you use a dependency? Deciding which packages (and how many) to depend on is a difficult and philosophical choice. Using functions from other packages can save you time and effort in development but it might make it more difficult to maintain your package. Some things you might want to consider before depending on a package are:

* How much of the functionality of the package do you want to use?
* Could you easily reproduce that functionality?
* How well maintained is the package?
* How often is it updated? Packages that change a lot are more likely to break your code.
* How many dependencies of it's own does that package have?
* Are you users likely to have the package installed already?

Packages like {ggplot2} are good choices for dependencies because they are well maintained, don't change too often, are commonly used and perform a single task so you are likely to use many of the functions.

Now if we change the `plot_colours` function to use the {ggplot2} namespace, we will get no more errors when we run `devtools::check()` but a note.

```{r}
#| label: plot_colours_ggplot2
#| eval: false

#' Plot colours
#'
#' Plot a vector of colours to see what they look like
#' 
#' @param colours Vector of colour to plot
#'
#' @return A ggplot2 object
#' @export
#'
#' @examples
#' shades <- make_shades("goldenrod", 5)
#' plot_colours(shades)
plot_colours <- function(colours) {
  plot_data <- data.frame(Colour = colours)
  
  ggplot2::ggplot(plot_data,
                  ggplot2::aes(x = .data$Colour, y = 1, fill = .data$Colour,
                               label = .data$Colour)) +
    ggplot2::geom_tile() +
    ggplot2::geom_text(angle = "90") +
    ggplot2::scale_fill_identity() +
    ggplot2::theme_void()
}
```

To deal with the note, we need to import the {rlang} package.

```{r}
#| label: use_rlang
#| eval: false

usethis::use_package("rlang")
```

Writing `rlang::.data` wouldn't be very attractive or readable. When we want to use a function in another package without `::` we need to explicitly import it. Just like when we exported our functions we do this using a {Roxygen} comment.


```{r}
#| label: plot_colours_import
#| eval: false

#' Plot colours
#'
#' Plot a vector of colours to see what they look like
#' 
#' @param colours Vector of colour to plot
#'
#' @return A ggplot2 object
#' @export
#'
#' @importFrom rlang .data
#' 
#' @examples
#' shades <- make_shades("goldenrod", 5)
#' plot_colours(shades)
plot_colours <- function(colours) {
  plot_data <- data.frame(Colour = colours)
  
  ggplot2::ggplot(plot_data,
                  ggplot2::aes(x = .data$Colour, y = 1, fill = .data$Colour,
                               label = .data$Colour)) +
    ggplot2::geom_tile() +
    ggplot2::geom_text(angle = "90") +
    ggplot2::scale_fill_identity() +
    ggplot2::theme_void()
}
```

If we used `rlang::.data` in multiple functions in our package it might make sense to only import it once. It doesn't matter where we put the `@importFrom` line (or how many times) it will still be added once to NAMESPACE. This means we can put all import in a central location. The advantage of this is that they only appear once and are all in one place but it makes it harder to know which of our functions have which imports and remove them if they are no longer needed. Which approach you take is up to you.

## Other documentation

### Package help file

Users can find out about our functions using `?function-name` but what if they want to find out about the package itself? There is some information in the DESCRIPTION but that can be hard to access. We can add a help file for the package using `usethis::use_package_doc()`.

```{r}
#| label: use_package_doc
#| eval: false

usethis::use_package_doc()
```

This command will create `'R/mypkg-package.R'`, which contains the following:

```{r}
#| label: mypkg-package.R
#| eval: false

#' @keywords internal
"_PACKAGE"

## usethis namespace: start
## usethis namespace: end
NULL
```

When we run `devtools::document()` a new .Rd file will be created and we can view the contents using `?mypkg`. The information here has been automatically pulled from the `DESCRIPTION` file so we only need to update it in one place.

### Vignettes

The documentation we have written so far explains how individual functions work in detail but it doesn't show what the package does as a whole. Vignettes are short tutorials that explain what the package is designed for and how different functions can be used together. There are different ways to write vignettes but usually they are R Markdown files. We can create a vignette with `usethis::use_vignette()`. There can be multiple vignettes but it is common practice to start with one that introduces the whole package.

```{r}
#| label: use_vignette
#| eval: false

usethis::use_vignette("mypkg")
```

This will create `vignettes/mypkg.Rmd` that you can use to create a tutorial for your package.

To see what the vignette looks like run `devtools::build_vignettes()`. Asking {devtools} to build the vignette rather than rendering it in another way (such as the Knit button in RStudio) makes sure that we are using the development version of the package rather than any version that is installed.

```{r}
#| label: build_vignette
#| eval: false

devtools::build_vignettes()
```

This creates `doc/` that contains the rendered vignette. If you want to use any other packages in your vignette that the package doesn't already depend on you need to add them as a suggested dependency.

### README

If you plan on sharing the source code rather than the built package it is useful to have a `README` file to explain what the package is, how to install and use it, how to contribute, etc. We can create a template with `usethis::use_readme_md()` (if we wanted to and R Markdown file with R code and output we might use `usethis::use_readme_rmd()` instead).

```{r}
#| label: use_readme_rmd
#| eval: false

usethis::use_readme_rmd()
```

This will create `README.Rmd` and will be used to render `README.md`.

There are the comments near the top that mention badges and you might have seen badges (or shields) on `README` files in code repositories before. There are several {usethis} functions for adding badges. For example we can mark this package as been at the experimental stage using `usethis::use_lifecycle_badge()`.

```{r}
#| label: use_lifecycle_badge
#| eval: false

usethis::use_lifecycle_badge("experimental")
```

### Package website

If you have a publicly available package it can be useful to have a website displaying the package documentation. It gives your users somewhere to go and helps your package appear in search results. Luckily this is easily achieved using the {pkgdown} package. If you have it installed you can set it up with {usethis}.

```{r}
#| label: use_pkgdown
#| eval: false

usethis::use_pkgdown()
```

## Versioning

Whenever you reach a milestone it is good to update the package version. Having a good versioning system is important when it comes to things like solving user issues. Version information is recorded in the `DESCRIPTION` file. This version number follows the format major.minor.patch.dev. The different parts of the version represent different things:

* major - A significant change to the package that would be expected to break users code. This is updated very rarely when the package has been redesigned in some way.
* minor - A minor version update means that new functionality has been added to the package. It might be new functions to improvements to existing functions that are compatible with most existing code.
* patch - Patch updates are bug fixes. They solve existing issues but don't do anything new.
* dev - Dev versions are used during development and this part is missing from release versions. For example you might use a dev version when you give someone a beta version to test. A package with a dev version can be expected to change rapidly or have undiscovered issues.

To increase the package version use `usethis::use_version()`.

```{r}
#| label: use_version
#| eval: false

usethis::use_version()
```
```
Current version is 0.0.0.9000.
What should the new version be? (0 to exit) 

1: major --> 1.0.0
2: minor --> 0.1.0
3: patch --> 0.0.1
4:   dev --> 0.0.0.9001
```

Whenever we update the package version we should record what changes have been made. We do this is a `NEWS.md` file.

```{r}
#| label: use_news_md
#| eval: false

usethis::use_news_md()
```

This will create 'NEWS.md' and we can modify this file to record what changes have been made.

## Building, installing, and releasing

If you want to start using your package in other projects the simplest thing to do is to run `devtools::install()`. This will install your package in the same way as any other package so that it can be loaded with `library()`. However this will only work on the computer you are developing the package on. If you want to share the package with other people (or other computers you work on) there are a few different options.

### Building

One way to share your package is to manually transfer it to somewhere else. But rather then copying the development directory what you should share is a pre-built package archive. Running `devtools::build()` will bundle up your package into a `.tar.gz` file without any of the extra bits required during development. This archive can then be transferred to wherever you need it and installed using `install.packages("mypkg.tar.gz", repos = NULL)` or `R CMD INSTALL mypkg.tar.gz` on the command line. While this is fine if you just want to share the package with yourself or a few people you know, it doesn't work if you want it to be available to the general community.

### CRAN

The most common repository for public R packages is the Comprehensive R Archive Network (CRAN). This is where packages are usually downloaded from when you use `install.packages()`. Compared to similar repositories for other programming languages, getting your package accepted to CRAN means meeting a series of requirements. While this makes the submission process more difficult it gives users confidence that your package is reliable and will work on multiple platforms. It also makes your package much easier to install for most users and makes it more discoverable.

See [Releasing to CRAN](https://r-pkgs.org/release.html) in the R packages book, [Getting your R package on CRAN ](https://kbroman.org/pkg_primer/pages/cran.html), and [Checklist for CRAN submissions](https://cran.r-project.org/web/packages/submission_checklist.html) for more information.

### Bioconductor

If your package is designed for analysing biological data you might want to submit it to [Bioconductor](https://www.bioconductor.org/) rather than CRAN. While Bioconductor has a smaller audience, it is more specialised and is often the first place researchers in the life sciences look. Building a Bioconductor package also means that you can take advantage of the extensive ecosystem of existing objects and packages for handling biological data types.

While there are lots of advantages to having your package on Bioconductor the coding style is slightly different to what is often used for CRAN packages. If you think you might want to submit your package to Bioconductor in the future check out <https://contributions.bioconductor.org/develop-overview.html>.

### rOpenSci

[rOpenSci](https://ropensci.org/) is not a package repository as such but an organisation that reviews and improves R packages. Packages that have been accepted by rOpenSci should meet a certain set of standards. By submitting your package to rOpenSci you will get it reviewed by experienced programmers who can offer suggestions on how to improve it. If you are accepted you will receive assistance with maintaining your package and it will be promoted by the organisation. Have a look at their [submission page](https://github.com/ropensci/software-review) for more details.

### Code sharing websites

Uploading your package to a code sharing website such as GitHub, Bitbucket or GitLab offers a good compromise between making your package available and going through an official submission process. This is a particularly good option for packages that are still in early development and are not ready to be submitted to one of the major repositories. Making your package available on one of these sites also gives it a central location for people to ask questions and submit issues. Code sharing websites are usually accessed through the Git version control system. If you are unfamiliar with using `git` on the command line there are functions in {usethis} that can run the commands for you from R.

To set up Git, first we need to configure our details.

```{r}
#| label: use_git_config
#| eval: false

usethis::use_git_config(user.name = "Your Name", user.email = "your@email.com")
```

The email address should be the same one you used to sign up to GitHub. Now we can set up Git in our package repository.

```{r}
#| label: use_git
#| eval: false

usethis::use_git()
```

The next step is to link the directory on your computer with a repository on GitHub. First we need to create a special access token. The following command will open a GitHub website.

```{r}
#| label: use_github
#| eval: false

usethis::use_github()
```

Click the "Generate token" button on the webpage and then copy the code on the next page. As it says you can only view this once so be careful to copy it now and don't close the page until you are finished. When you have the code follow the rest of the instructions from usethis.

```{r}
#| label: edit_r_environ
#| eval: false

usethis::edit_r_environ()
```

Edit the file to look something like this (with your code).

```
GITHUB_PAT=YOUR_CODE_GOES_HERE
```

Save it then restart R by clicking the Session menu and selecting Restart R.

Copying that code and adding it to your `.Renviron` gives R on the computer you are using access to your GitHub repositories. If you move to a new computer you will need to do this again. Now that we have access to GitHub we can create a repository for our packages.

```{r}
#| label: use_github_after_setup
#| eval: false

usethis::use_github()
```

Respond to the prompts from {usethis} about the method for connecting to GitHub and the title and description for the repository. When everything is done a website should open with your new package repository. Another thing this function does is add some extra information to the description that lets people know where to find your new website.

```
URL: https://github.com/user/mypkg
BugReports: https://github.com/user/mypkg/issues
```

Now that your package is on the internet anyone can install it using the `install_github()` function in the {remotes} package (which should already be installed as a dependency of {devtools}). All you need to give it is the name of the user and repository.

```{r}
#| label: install_github
#| eval: false

remotes::install_github("user/mypkg")
```

If you are familiar with Git you can install from a particular branch, tag or commit by adding that after @.

```{r}
#| label: install_github_specific
#| eval: false

remotes::install_github("user/mypkg@branch_name")
remotes::install_github("user/mypkg@tag_id")
remotes::install_github("user/mypkg@commit_sha")
```

After you make improvements to your package you will probably want to update the version that is online. To do this you need to learn a bit more about git. Jenny Bryan's "[Happy Git with R](https://happygitwithr.com/) tutorial is a great place to get started but the (very) quick steps in RStudio are:

1. Open the Git pane (it will be with Environment, History etc.)
2. Click the check box next to each of the listed files
3. Click the Commit button
4. Enter a message in the window that opens and click the Commit button
5. Click the Push button with the green up arrow

Refresh the GitHub repository website and you should see the changes you have made.

## Advanced topics

Most of the following topics are covered in Hadley Wickham's [Advanced R book](https://adv-r.hadley.nz/).

### Including datasets

It can be useful to include small datasets in your R package which can be used for testing and examples in your vignettes. You may also have reference data that is required for your package to function. If you already have the data as an object in R it is easy to add it to your package with `usethis::use_data()`. The `usethis::use_data_raw()` function can be used to write a script that reads a raw data file, manipulates it in some way and adds it to the package with `usethis::use_data()`. This is useful for keeping a record of what you have done to the data and updating the processing or dataset if necessary.

See the [Data section of R packages](https://r-pkgs.org/data.html) for more details about including data in your package.

### Designing objects

If you work with data types that don't easily fit into a table or matrix you may find it convenient to design specific objects to hold them. Objects can also be useful for holding the output of functions such as those that fit models or perform tests. R has several different object systems.

The S3 system is the simplest and probably the most commonly used. Packages in the Bioconductor ecosystem make use of the more formal S4 system. If you want to learn more about designing R objects a good place to get started is the [Object-oriented programming chapter](https://adv-r.hadley.nz/oo.html) of Hadley Wickham's Advanced R book. Other useful guides include Nicholas Tierney's [A Simple Guide to S3 Methods](https://arxiv.org/abs/1608.07161) and Stuart Lee's [S4: a short guide for the perplexed](https://stuartlee.org/2019/07/09/s4-short-guide/).

### Integrating other languages

If software for completing a task already exists but is in another language it might make sense to write an R package that provides an interface to the existing implementation rather than re-implementing it from scratch. Here are some of the R packages that help you integrate code from other languages:

* Rcpp (C++) <https://www.rcpp.org/>
* reticulate (Python) <https://rstudio.github.io/reticulate/>
* RStan (Stan) <https://mc-stan.org/users/interfaces/rstan>
* rJava (Java) <https://www.rforge.net/rJava/>

Another common reason to include code from another language is to improve performance. While it is often possible to make code faster by re-considering how things are done within R, sometimes there is no alternative. The {Rcpp} package makes it very easy to write snippets of C++ code that is called from R. Depending on what you are doing moving even very small bits of code to C++ can have big impacts on performance. Using {Rcpp} can also provide access to existing C libraries for specialised tasks.

The [Rewriting R code in C++](https://adv-r.hadley.nz/rcpp.html) section of Advanced R explains when and how to use {Rcpp}. You can find other resources including a gallery of examples on the [official Rcpp website](https://www.rcpp.org/).

### Metaprogramming

Metaprogramming refers to code that reads and modifies other code. This may seem like an obscure topic but it is important in R because of its relationship to non-standard evaluation (another fairly obscure topic). You may not have heard of non-standard evaluation before but it is likely you have used it. This is what happens whenever you provide a function with a bare name instead of a string or a variable. Metaprogramming particularly becomes relevant to package development if you want to have functions that make use of packages in the Tidyverse such as {dplyr}, {tidyr}, and {purrr}. The [Metaprogramming](https://adv-r.hadley.nz/metaprogramming.html) chapter of Advanced R covers the topic in more detail.

# Session information

```{r session_info}
sessionInfo()
```
