---
title: "R package workshop"
format:
  gfm:
    toc: true
date: '`r Sys.Date()`'
date-format: iso
engine: knitr
---

```{r load_libraries, include=FALSE}
library(tidyverse)
```

# Introduction

This [R package workshop](https://combine-australia.github.io/r-pkg-dev/) was created by COMBINE, an association for Australian students in bioinformatics, computational biology, and related fields. It is licensed under a [Creative Commons Attribution 4.0 International (CC BY 4.0) license]((https://creativecommons.org/licenses/by/4.0/)).

An R package is a collection of functions that are bundled together in a way that makes it easy to share. Usually these functions are designed to work together to complete a specific task such as analysing a particular kind of data. Packages are the best way to distribute code and documentation; even if you never intend to share your package it is useful to have a place to store your commonly used functions. It's a bit more effort now but it will save you a lot of time in the long run.

This workshop teaches a modern package development workflow that makes use of packages designed to help with writing packages. The two main packages are {devtools} and {usethis}:

* {devtools} contains functions that will help with development tasks such as checking, building and installing packages.
* {usethis} contains a range of templates and handy functions for making life easier, many of which were originally in {devtools}.

Other packages used in this workshop include:

* {roxygen2} for function documentation
* {testthat} for writing unit tests
* {knitr} for building vignettes

Install the required packages for this workshop.

```{r}
#| label: required_packages
#| eval: false

pkgs <- c("devtools", "usethis", "roxygen2", "testthat", "knitr", "ggplot2", "rlang")
install.packages(pkgs, Ncpus = 2)
```

Some stages of the package development process can require other programs to be installed on your computer. To check that you have everything you need run the following:

```{r}
#| label: check_build_tools

pkgbuild::check_build_tools()
```

# Session information

```{r session_info}
sessionInfo()
```
