---
title: "R package workshop"
format:
  gfm:
    toc: true
date: '`r Sys.Date()`'
date-format: iso
engine: knitr
---

```{r load_libraries, include=FALSE}
library(tidyverse)
```

# TL;DR

Install packages, check whether the system is ready to build packages, initialise package, add license, create file to hold functions, and check for issues.

```{r}
#| label: tl_dr
#| eval: false

pkgs <- c("devtools", "usethis", "roxygen2", "testthat", "knitr", "ggplot2", "rlang")
install.packages(pkgs, Ncpus = 2)
pkgbuild::check_build_tools()

usethis::create_package(path = '/tmp/mypkg')
usethis::use_mit_license("Dave Tang")
usethis::use_r("colours")
devtools::document()
usethis::use_testthat()
usethis::use_test("colours")
devtools::test()
devtools::check()
```

# Introduction

This [R package workshop](https://combine-australia.github.io/r-pkg-dev/) was created by COMBINE, an association for Australian students in bioinformatics, computational biology, and related fields. It is licensed under a [Creative Commons Attribution 4.0 International (CC BY 4.0) license](https://creativecommons.org/licenses/by/4.0/).

An R package is a collection of functions that are bundled together in a way that makes it easy to share. Usually these functions are designed to work together to complete a specific task such as analysing a particular kind of data. Packages are the best way to distribute code and documentation; even if you never intend to share your package it is useful to have a place to store your commonly used functions. It's a bit more effort now but it will save you a lot of time in the long run.

This workshop teaches a modern package development workflow that makes use of packages designed to help with writing packages. The two main packages are {devtools} and {usethis}:

* {devtools} contains functions that will help with development tasks such as checking, building and installing packages.
* {usethis} contains a range of templates and handy functions for making life easier, many of which were originally in {devtools}.

Other packages used in this workshop include:

* {roxygen2} for function documentation
* {testthat} for writing unit tests
* {knitr} for building vignettes

Install the required packages for this workshop.

```{r}
#| label: required_packages
#| eval: false

pkgs <- c("devtools", "usethis", "roxygen2", "testthat", "knitr", "ggplot2", "rlang")
install.packages(pkgs, Ncpus = 2)
```

Some stages of the package development process can require other programs to be installed on your computer. To check that you have everything you need run the following:

```{r}
#| label: check_build_tools

pkgbuild::check_build_tools()
```

# Getting started

To begin building a package, we need a package name; names can only consist of letters, numbers and dots (.) and must start with a letter. While all of these are allowed it is generally best to stick to just lowercase letters. Having a mix of lower and upper case letters can be hard for users to remember (is it RColorBrewer or Rcolorbrewer or rcolorbrewer?). For this workshop, we are going to use {mypkg}.

To create a template for our package, use `usethis::create_package()`.

```{r}
#| label: create_package
#| eval: false
 
usethis::create_package(path = '/tmp/mypkg')
```

Several files get created.

```{bash}
ls -R /tmp/mypkg
```

* `DESCRIPTION` contains the metadata for your package.
* `NAMESPACE` describes the functions in our package; it should not be manually edited.
* `mypkg.Rproj` is a RStudio project file.
* `R` is the directory that will hold all our R code.

These files are the minimal amount that is required for a package but we will create other files as we go along. Some other useful (hidden) files have also been created by {usethis}:

* `.gitignore` is useful if you use `git` for version control.
* `.Rbuildignore` is used to mark files that are in the directory but aren’t really part of the package and shouldn't be included when we build it. Most of the time you won't need to worry about this as {usethis} will edit it for you.

## DESCRIPTION

The `DESCRIPTION` file is one of the most important parts of a package. It contains all the metadata about the package, things like what the package is called, what version it is, a description, who the authors are, what other packages it depends on, etc.

```{bash}
cat /tmp/mypkg/DESCRIPTION
```

* The `Title` should be a single line in Title Case that explains what your package is.
* The `Description` is a paragraph which goes into a bit more detail about what the package does.

The `Authors@R` field shows an example of how to define an author. You can see that the example person has been assigned the author ("aut") and creator ("cre") roles. There must be at least one author and one creator for every package (they can be the same person) and the creator must have an email address. Other roles include:

* `cre`: the creator or maintainer of the package, the person who should be contacted when there are problems
* `aut`: authors, people who have made significant contributions to the package
* `ctb`: contributors, people who have made smaller contributions
* `cph`: copyright holder, useful if this is someone other than the creator (such as their employer)

## License

The software license describes how our code can be used and without one people must assume that it can't be used at all! For this example we will use the MIT license which basically says the code can be used for any purpose and doesn't come with any warranties. There are templates for some of the most common licenses included in {usethis}.

```{r}
#| label: use_mit_license
#| eval: false
 
usethis::use_mit_license("Dave Tang")
```

## Functions

Use `usethis::use_r()` to create an empty file in `R` to hold our function about colours.

```{r}
#| label: use_r
#| eval: false

usethis::use_r("colours")
```

There are no rules about how to organise your functions into different files but in general similar functions should be grouped together into the same file with a clear name. Putting all functions into a single file is not ideal but neither is having a separate file for each function. A good rule of thumb is that if you are finding it hard to locate a function you might need to move it to a new file.

The example for this workshop is a function that takes the red, green and blue values for a colour and returns a given number of shades. Save the function below into `R/colours.R`.

```{r}
#| label: make_shades

make_shades <- function(colour, n, lighter = TRUE) {
  # Convert the colour to RGB
  colour_rgb <- grDevices::col2rgb(colour)[, 1]
  
  # Decide if we are heading towards white or black
  if (lighter) {
    end <- 255
  } else {
    end <- 0
  }
  
  # Calculate the red, green and blue for the shades
  # we calculate one extra point to avoid pure white/black
  red   <- seq(colour_rgb[1], end, length.out = n + 1)[1:n]
  green <- seq(colour_rgb[2], end, length.out = n + 1)[1:n]
  blue  <- seq(colour_rgb[3], end, length.out = n + 1)[1:n]
  
  # Convert the RGB values to hex codes
  shades <- grDevices::rgb(red, green, blue, maxColorValue = 255)
  
  return(shades)
}
```

Usually when we write a new function we load it by copying the code to the console or sourcing the R file. When we are developing a package we want to try and keep our environment empty so that we can be sure we are only working with objects inside the package. Instead we can load functions using `devtools::load_all()`.

## Checking your package

The `devtools::check()` function runs a series of checks to make sure your package is working correctly. It is highly recommended that you run `devtools::check()` often and follow its advice to fix any problems.

```{r}
#| label: devtools_check
#| eval: false

devtools::check()
```

You will see all the different types of checks that {devtools} has run but the most important section is at the end where it tells you how many errors, warnings, and notes there are.

* Errors happen when your code has broken and failed one of the checks. If errors are not fixed your package will not work correctly.
* Warnings are slightly less serious but should also be addressed. Your package will probably work without fixing them but it is highly advised that you do.
* Notes are advice rather than problems. It can be up to you whether or not to address them but there is usually a good reason to.

Often the failed checks come with hints about how to fix them but sometimes they can be hard to understand.

## Documenting our function

The {roxygen2} package makes it easy to document our function by writing some special comments at the start of our function. This has the extra advantage of keeping the documentation with the code which make it easier to keep it up to date.

If you are using RStudio, open the "Code" menu and select "Insert Roxygen skeleton", which will create the template shown below. The parameters were automatically generated.

```{r}
#| label: roxygen_skeleton

#' Title
#'
#' @param colour 
#' @param n 
#' @param lighter 
#'
#' @return
#' @export
#'
#' @examples
```

Roxygen comments all start with `#'`. The first line is the title of the function followed by a blank line. Next is a paragraph that provides a more detailed description of the function.

The next section describes the parameters for the function marked by the `@param` field.

The next field is `@return`, which is used to describe what the function returns. This is usually fairly short but you should provide enough detail to make sure that the user knows what they are getting back.

The next field is `@export` and is a bit different because it doesn't add documentation to the help file but instead modifies the `NAMESPACE` file. Adding `@export` tells {Roxygen} that this is a function that we want to be available to the user. When we build the documentation, {Roxygen} will then add the correct information to the `NAMESPACE` file. If we had an internal function that wasn't meant to be used by the user we would leave out `@export`.

The last field in the skeleton is `@examples` and is where we put some short examples showing how the function can be used. These will be placed in the help file and can be run using `example("function")`.

```{r}
#| label: roxygen_skeleton_filled

#' Make shades
#'
#' Given a colour make n lighter or darker shades
#' 
#' @param colour The colour to make shades of
#' @param n The number of shades to make
#' @param lighter Whether to make lighter (TRUE) or darker (FALSE) shades
#'
#' @return A vector of n colour hex codes
#' @export
#'
#' @examples
#' # Five lighter shades
#' make_shades("goldenrod", 5)
#' # Five darker shades
#' make_shades("goldenrod", 5, lighter = FALSE)
```

There are other useful fields such as `@author` (specify the function author), `@references` (any associated references) and `@seealso` (links to related functions).

Use `devtools::document()` to build the documentation.

```{r}
#| label: document
#| eval: false

devtools::document()
```

## Testing

Unit tests are checks to make sure that a function works in the way that we expect. The examples in the documentation act like informal unit tests because they are run as part of the checking process but it is better to have something more rigorous.

One approach to writing unit tests is what is known as "test driven development". The idea is to write the tests before you write a function. This way you know exactly what a function is supposed to do and what problems there might be. While this is a good principal it can take a lot of advance planning.

A more common approach could be called "bug-driven testing". For this approach whenever we come across a bug we write a test for it before we fix it, that way the same bug should never happen a again. When combined with some tests for obvious problems this is a good compromise between testing for every possible outcome and not testing at all.

For example, let's see what happens when we ask `make_shades()` for a negative number of shades.

```{r}
#| label: make_shades_error
#| error: true

make_shades("goldenrod", -1)
```

Before we make any changes to the function let's design some tests to make sure we get what we expect. There are a few ways to write unit tests for R packages but we are going to use the {testthat} package. We can set everything up with {usethis}.

```{r}
#| label: use_test_that
#| eval: false

usethis::use_testthat()
```

A `tests/` directory will be created to hold all our tests. There is also a `tests/testthat.R` file which looks like this:

```{r}
#| label: test_that_r_script
#| eval: false

# This file is part of the standard setup for testthat.
# It is recommended that you do not modify it.
#
# Where should you do additional test configuration?
# Learn more about the roles of various files in:
# * https://r-pkgs.org/testing-design.html#sec-tests-files-overview
# * https://testthat.r-lib.org/articles/special-files.html

library(testthat)
library(mypkg)

test_check("mypkg")
```

The above script makes sure that our tests are run with `devtools::check()`. To open a new test file use `usethis::use_test()`.

```{r}
#| label: use_test_colours
#| eval: false

usethis::use_test("colours")
```

`use_test("colours")` creates `tests/testthat/test-colours.R`, which matches up with the R files but with `test-` as the prefix. This test file comes with a small example that shows how to use {testthat}.

```{r}
#| label: test_that_example
#| eval: false

test_that("multiplication works", {
  expect_equal(2 * 2, 4)
})
```

Each set of tests starts with the `test_that()` function. This function has two arguments, a description and the code with the tests that we want to run. Inside the code section we see an expect function. This function also has two parts, the thing we want to test and what we expect it to be. There are different functions for different types of expectations.

For our test we want to use the `expect_error()` function, because that is what we expect.

```{r}
#| label: test_that_n_is_at_least_one
#| eval: false

test_that("n is at least 1", {
  expect_error(make_shades("goldenrod", -1),
               "n must be at least 1")
  expect_error(make_shades("goldenrod", 0),
               "n must be at least 1")
})
```

To run our tests we use `devtools::test()`. We will see that both of our tests failed. The first test fails because the error message is wrong and the second one because there is no error. Now that we have some tests and we know they check the right things we can modify our function to check the value of `n` and give the correct error.

Let's add some code to check the value of n. We will update the documentation as well so the user knows what values can be used.

```{r}
#| label: make_shades_check_n

#' Make shades
#'
#' Given a colour make \code{n} lighter or darker shades
#'
#' @param colour The colour to make shades of
#' @param n The number of shades to make, at least 1
#' @param lighter Whether to make lighter (\code{TRUE}) or darker (\code{FALSE})
#' shades
#'
#' @return A vector of \code{n} colour hex codes
#' @export
#'
#' @examples
#' # Five lighter shades
#' make_shades("goldenrod", 5)
#' # Five darker shades
#' make_shades("goldenrod", 5, lighter = FALSE)
make_shades <- function(colour, n, lighter = TRUE) {
  
  # Check the value of n
  if (n < 1) {
    stop("n must be at least 1")
  }
  
  # Convert the colour to RGB
  colour_rgb <- grDevices::col2rgb(colour)[, 1]
  
  # Decide if we are heading towards white or black
  if (lighter) {
    end <- 255
  } else {
    end <- 0
  }
  
  # Calculate the red, green and blue for the shades
  # we calculate one extra point to avoid pure white/black
  red <- seq(colour_rgb[1], end, length.out = n + 1)[1:n]
  green <- seq(colour_rgb[2], end, length.out = n + 1)[1:n]
  blue <- seq(colour_rgb[3], end, length.out = n + 1)[1:n]
  
  # Convert the RGB values to hex codes
  shades <- grDevices::rgb(red, green, blue, maxColorValue = 255)
  
  return(shades)
}
```

These kinds of checks for parameter inputs are an important part of a function that is going to be used by other people (or future you). They make sure that all the input is correct before the function tries to do anything and avoids confusing error messages. However they can be fiddly and repetitive to write. If you find yourself writing lots of these checks two packages that can make life easier by providing functions to do it for you are {checkmate} and {assertthat}.

In the code above we have used the `stop()` function to raise an error. If we wanted to give a warning we would use `warning()` and if just wanted to give some information to the user we would use `message()`. Using `message()` instead of `print()` or `cat()` is important because it means the user can hide the messages using `suppressMessages()` (or `suppressWarnings()` for warnings). Now if we try our tests again they should pass.

If you want to see what parts of your code need testing you can run the `devtools::test_coverage()` function (you will need to install the {DT} and {covr} packages first). This function uses the {covr} package to make a report showing which lines of your code are covered by tests.

# Session information

```{r session_info}
sessionInfo()
```
